The Proofs

### 2024-07-09 proofs/tex/decl-subtyping-lemmas.tex
edit and prepare for jfp

      \end{itemize}
    \item [$-_{L}$] Suppose $[[iM]] \neq [[∀]]\dots$. Let us prove both directions.
      \begin{itemize}
        \item [$\Rightarrow$] Let us assume $[[Γ ⊢ ∀pas.iN ≤ iM]]$.
          then if $[[pas = ·]]$, $[[Γ ⊢ iN ≤ iM]]$ holds immediately.
          Otherwise, let us decompose  $[[iN]]$ as $[[∀pas'.iN']]$ where 
          $[[iN']]$ does not start with $[[∀]]$.
          Then $[[Γ ⊢ ∀pas.∀pas'.iN' ≤ iM']]$ is inferred by
          \ruleref{\ottdruleDOneForallLabel},
          and by inversion, 
          there exist $[[Γ ⊢ iPs,iPs']]$ 
          such that $[[Γ ⊢ [iPs/pas][iPs'/pas']iN' ≤ iM']]$ 
          (the decomposition of substitutions is possible since $[[{pas} ∩ Γ = ∅]]$).
          Then by \ruleref{\ottdruleDOneForallLabel} again,
          $[[Γ ⊢ ∀pas'.[iPs'/pas']iN' ≤ iM']]$ (notice that $[[ [iPs'/pas']iN' ]]$ cannot
          start with $[[∀]]$).
        \item [$\Leftarrow$] Let us assume 
          $[[Γ ⊢ [iPs/pas]iN ≤ iM]]$ for some $[[Γ ⊢ iPs]]$.
          let us decompose $[[iN]]$ as $[[∀pas'.iN']]$ where $[[iN']]$ does not start with $[[∀]]$.
          Then $[[Γ ⊢ [iPs/pas]∀pas'.iN' ≤ iM']]$ or, equivalently,
          $[[Γ ⊢ ∀pas'.[iPs/pas]iN' ≤ iM']]$ is inferred by \ruleref{\ottdruleDOneForallLabel}
          (notice that $[[ [iPs/pas]iN' ]]$ cannot start with $[[∀]]$).
          By inversion, there exist $[[Γ ⊢ iPs']]$ such that 
          $[[Γ ⊢ [iPs'/pas'][iPs/pas]iN' ≤ iM']]$. Since $[[pas']]$ is disjoint
          from the free variables of $[[iPs]]$ and from $[[pas]]$, the composition of 
          $[[iPs'/pas']]$ and $[[iPs/pas]]$ can be joined into a single substitution
          well-formed in $[[Γ]]$. Then by \ruleref{\ottdruleDOneForallLabel} again,
          $[[Γ ⊢ ∀pas.iN ≤ iM]]$.
      \end{itemize}
      \item [$+$] The positive cases are proved symmetrically.
  \end{itemize}
\end{proof}

>>>>

      \end{itemize}
    \item [$-_{L}$] Suppose $[[iM]] \neq [[∀]]\dots$. Let us prove both directions.
      \begin{itemize}
        \item [$\Rightarrow$] Let us assume $[[Γ ⊢ ∀pas.iN ≤ iM]]$.
          then if $[[pas = ·]]$, $[[Γ ⊢ iN ≤ iM]]$ holds immediately.
          Otherwise, let us decompose  $[[iN]]$ as $[[∀pas'.iN']]$ where 
          $[[iN']]$ does not start with $[[∀]]$.
          Then $[[Γ ⊢ ∀pas.∀pas'.iN' ≤ iM']]$ is inferred by
          \ruleref{\ottdruleDOneForallLabel},
          and by inversion, 
          there exist $[[Γ ⊢ iPs]]$ and $[[Γ ⊢ iPs']]$ 
          such that $[[Γ ⊢ [iPs/pas][iPs'/pas']iN' ≤ iM']]$ 
          (the decomposition of substitutions is possible since $[[{pas} ∩ Γ = ∅]]$).
          Then by \ruleref{\ottdruleDOneForallLabel} again,
          $[[Γ ⊢ ∀pas'.[iPs'/pas']iN' ≤ iM']]$ (notice that $[[ [iPs'/pas']iN' ]]$ cannot
          start with $[[∀]]$).
        \item [$\Leftarrow$] Let us assume 
          $[[Γ ⊢ [iPs/pas]iN ≤ iM]]$ for some $[[Γ ⊢ iPs]]$.
          let us decompose $[[iN]]$ as $[[∀pas'.iN']]$ where $[[iN']]$ does not start with $[[∀]]$.
          Then $[[Γ ⊢ [iPs/pas]∀pas'.iN' ≤ iM']]$ or, equivalently,
          $[[Γ ⊢ ∀pas'.[iPs/pas]iN' ≤ iM']]$ is inferred by \ruleref{\ottdruleDOneForallLabel}
          (notice that $[[ [iPs/pas]iN' ]]$ cannot start with $[[∀]]$).
          By inversion, there exist $[[Γ ⊢ iPs']]$ such that 
          $[[Γ ⊢ [iPs'/pas'][iPs/pas]iN' ≤ iM']]$. Since $[[pas']]$ is disjoint
          from the free variables of $[[iPs]]$ and from $[[pas]]$, the composition of 
          $[[iPs'/pas']]$ and $[[iPs/pas]]$ can be joined into a single substitution
          well-formed in $[[Γ]]$. Then by \ruleref{\ottdruleDOneForallLabel} again,
          $[[Γ ⊢ ∀pas.iN ≤ iM]]$.
      \end{itemize}
      \item [$+$] The positive cases are proved symmetrically.
  \end{itemize}
\end{proof}

### 2024-07-09 proofs/tex/subst-lemmas.tex
edit and prepare for jfp

\lemmaSubstRestrFv*
\begin{proof}
  First, we strengthen the statement by saying that
  one can restrict the substitution to an arbitrary
  superset of the free variables of the substitution subject:
 \begin{itemize}
    \item[$+$] $[[ [σ]iP ]] = [[ [σ|varset]iP ]]$, for any $[[varset]] \supseteq [[fv iP]]$, and
    \item[$-$] $[[ [σ]iN ]] = [[ [σ|varset]iN ]]$, for any $[[varset]] \supseteq [[fv iN]]$.
  \end{itemize}
  Then the proof is a straightforward induction on $[[iP]]$ and mutually, on $[[iN]]$.
  For the base cases: 
  \begin{caseof}
    \item $[[iN = α⁻]]$\\
      Then $[[ [σ]α⁻ ]] = [[σ|varset]] ([[α⁻]])$ by definition, 
      since $[[α⁻]] \in [[fv α⁻ ⊆ varset]]$.
    \item $[[iN = iP → iM]]$\\
      Then $[[ [σ](iP → iM) ]] = [[ [σ]iP → [σ]iM ]]$ by definition.
      Since $[[fv iP]] \subseteq [[fv(iP → iM)]] \subseteq [[varset]]$, 
      the induction hypothesis is applicable to $[[ [σ]iP ]]$: 
      $[[ [σ]iP ]] = [[ [σ|varset]iP ]]$. 
      Analogously, and $[[ [σ]iM ]] = [[ [σ|varset]iM ]]$.
      Then 
      $[[ [σ](iP → iM) ]] = [[ [σ|varset]iP → [σ|varset]iM ]]
       = [[ [σ|varset](iP → iM) ]]$.
    \item $[[iN = ↑iP]]$ is proved analogously to the previous case.
    \item $[[iN = ∀pas.iM ]]$ (where $[[pas]]$ is not empty)\\
      Then $[[ [σ]∀pas.iM ]] = [[ ∀pas.[σ]iM ]]$ by definition.
      Let us assume $[[pas]]$ are fresh variables, 
      it means that $[[σ]]([[α± ]]) = [[α± ]]$ for any 
      $[[α± ]] \in [[pas]]$, and thus, 
      $[[ [σ|varset] ]] = [[ [σ|(varset ∪ {pas})] ]]$
      immediately from the definition.

      Since $[[varset ⊆ fv(∀pas.iM)]] = [[fv iM \ {pas}]]$,
      $[[varset ∪ {pas} ⊆ fv(iM)]]$.
      Then by the induction hypothesis, 
      $[[ [σ]iM =  [σ|(varset ∪ {pas})]iM ]]$.
      Finally, 
      $[[ [σ]∀pas.iM ]] = [[ ∀pas.[σ|(varset ∪ {pas})]iM ]] = 
      [[ ∀pas.[σ|varset]iM ]] = [[ [σ|varset]∀pas.iM ]]$.
    \item The positive cases are proves symmetrically. 
  \end{caseof}
\end{proof}

>>>>

\lemmaSubstRestrFv*
\begin{proof}
  First, we strengthen the statement by saying that
  one can restrict the substitution to an arbitrary
  superset of the free variables of the substitution subject:
 \begin{itemize}
    \item[$+$] $[[ [σ]iP ]] = [[ [σ|varset]iP ]]$, for any $[[varset]] \supseteq [[fv iP]]$, and
    \item[$-$] $[[ [σ]iN ]] = [[ [σ|varset]iN ]]$, for any $[[varset]] \supseteq [[fv iN]]$.
  \end{itemize}
  Then the proof is a straightforward induction on $[[iP]]$ and mutually, on $[[iN]]$.
  For the base cases: 
  \begin{caseof}
    \item $[[iN = α⁻]]$\\
      Then $[[ [σ]α⁻ ]] = [[σ|varset]] ([[α⁻]])$ by definition, 
      since $[[α⁻]] \in [[fv α⁻ ⊆ varset]]$.
    \item $[[iN = iP → iM]]$\\
      Then $[[ [σ](iP → iM) ]] = [[ [σ]iP → [σ]iM ]]$ by definition.
      Since $[[fv iP]] \subseteq [[fv(iP → iM)]] \subseteq [[varset]]$, 
      the induction hypothesis is applicable to $[[ [σ]iP ]]$: 
      $[[ [σ]iP ]] = [[ [σ|varset]iP ]]$. 
      Analogously, and $[[ [σ]iM ]] = [[ [σ|varset]iM ]]$.
      Then 
      $[[ [σ](iP → iM) ]] = [[ [σ|varset]iP → [σ|varset]iM ]]
       = [[ [σ|varset](iP → iM) ]]$.
    \item $[[iN = ↑iP]]$ is proved analogously to the previous case.
    \item $[[iN = ∀pas.iM ]]$ (where $[[pas]]$ is not empty)\\
      Then $[[ [σ]∀pas.iM ]] = [[ ∀pas.[σ]iM ]]$ by definition.
      Let us assume $[[pas]]$ are fresh variables, 
      it means that $[[σ]]([[α± ]]) = [[α± ]]$ for any 
      $[[α± ]] \in [[pas]]$, and thus, 
      $[[ σ|varset ]] = [[ σ|(varset ∪ {pas}) ]]$
      immediately from the definition.

      Since $[[varset ⊆ fv(∀pas.iM)]] = [[fv iM \ {pas}]]$,
      $[[varset ∪ {pas} ⊆ fv(iM)]]$.
      Then by the induction hypothesis, 
      $[[ [σ]iM =  [σ|(varset ∪ {pas})]iM ]]$.
      Finally, 
      $[[ [σ]∀pas.iM ]] = [[ ∀pas.[σ|(varset ∪ {pas})]iM ]] = 
      [[ ∀pas.[σ|varset]iM ]] = [[ [σ|varset]∀pas.iM ]]$.
    \item The positive cases are proven symmetrically. 
  \end{caseof}
\end{proof}

### 2024-07-09 proofs/tex/translation-proofs.tex
edit and prepare for jfp

    \item \ruleref{\ottdruleELTAppLabel}
      By inversion of this rule, we have
      \begin{itemize}
        \item $[[fΓ ; fΦ ⊢ ft : ∀a.fT  ⤳  c]]$, and thus, 
          $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|∀a.fT|+]]$ by the induction hypothesis; and
        \item $[[fΓ ⊢ fA]]$, which implies $[[+|fΓ|+ ⊢ +|fA|+]]$ by \cref{lemma:typ-pol-wf-completeness}.
      \end{itemize}
      We wish to infer
      $[[+|fΓ|+ ; +|fΦ|+ ⊢ let f : +|∀a.fT|+ = c; let y : +|[fA/a]fT|+ = f(); return y : ↑+|[fA/a]fT|+]]$.
      First, let us rewrite the polarized types by definition and \cref{lemma:pol-subst}:
      $[[+|fΓ|+ ; +|fΦ|+ ⊢ let f : ↓∀α⁺.↑+|fT|+ = c; let y : [+|fA|+/α⁺]+|fT|+ = f(); return y : ↑[+|fA|+/α⁺]+|fT|+]]$.
      Then we prove this judgment gradually by applying the corresponding rules of \fexists:
      \begin{enumerate}
        \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+, y : [+|fA|+/α⁺]+|fT|+ ⊢ return y : ↑[+|fA|+/α⁺]+|fT|+]]$
          holds by \ruleref{\ottdruleDTVarLabel} and \ruleref{\ottdruleDTReturnLabel}.
        \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ let y : [+|fA|+/α⁺]+|fT|+ = f(); return y : ↑[+|fA|+/α⁺]+|fT|+]]$
          holds by \ruleref{\ottdruleDTAppLetAnnLabel}. Let us show that the 
          required premises hold. 
          \begin{enumerate}
            \item $[[+|fΓ|+ ⊢ [+|fA|+/α⁺]+|fT|+]]$ holds since $[[+|fΓ|+ ⊢ +|fA|+]]$ and $[[+|fΓ|+ ⊢ +|fT|+]]$ by \cref{lemma:wf-subst}.
            \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ f : ↓∀α⁺.↑+|fT|+]]$ by \ruleref{\ottdruleDTVarLabel}.
            \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ ∀α⁺.↑+|fT|+ ● · ⇒> ∀α⁺.↑+|fT|+]]$ by \ruleref{\ottdruleDTEmptyAppLabel}.
            \item $[[+|fΓ|+  ⊢ ∀α⁺.↑+|fT|+ ≤ ↑[+|fA|+/α⁺]+|fT|+ ]]$ by \ruleref{\ottdruleDOneForallLabel}:
              $[[+|fΓ|+  ⊢ +|fA|+/α⁺ : {α⁺}]]$ since $[[+|fΓ|+  ⊢ +|fA|+]]$, as noted above.
            \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+, y : [+|fA|+/α⁺]+|fT|+ ⊢ return y : ↑[+|fA|+/α⁺]+|fT|+]]$ as noted above.
          \end{enumerate}
        \item $[[+|fΓ|+ ; +|fΦ|+  ⊢ let f : ↓∀α⁺.↑+|fT|+ = c; let y : [+|fA|+/α⁺]+|fT|+ = f(); return y : ↑[+|fA|+/α⁺]+|fT|+]]$
          holds by \ruleref{\ottdruleDTCVarLetLabel}. Let us show that the
          required premises hold.
          \begin{enumerate}
            \item $[[+|fΓ|+ ⊢ ↓∀α⁺.↑+|fT|+]]$ by \cref{lemma:typ-pol-wf-completeness}.
            \item $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑↓∀α⁺.↑+|fT|+]]$ by the induction hypothesis, as noted above.
            \item $[[+|fΓ|+  ⊢ ↑↓∀α⁺.↑+|fT|+ ≤ ↑↓∀α⁺.↑+|fT|+]]$ by reflexivity (\cref{lemma:subtyping-reflexivity}).
            \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ let y : [+|fA|+/α⁺]+|fT|+ = f(); return y : ↑[+|fA|+/α⁺]+|fT|+]]$ as noted above.
          \end{enumerate}
      \end{enumerate}
  \end{caseof}
\end{proof}

>>>>

    \item \ruleref{\ottdruleELTAppLabel}
      By inversion of this rule, we have
      \begin{itemize}
        \item $[[fΓ ; fΦ ⊢ ft : ∀a.fT  ⤳  c]]$, and thus, 
          $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|∀a.fT|+]]$ by the induction hypothesis; and
        \item $[[fΓ ⊢ fA]]$, which implies $[[+|fΓ|+ ⊢ +|fA|+]]$ by \cref{lemma:typ-pol-wf-completeness}.
      \end{itemize}
      We wish to infer
      $[[+|fΓ|+ ; +|fΦ|+ ⊢ let f : +|∀a.fT|+ = c; let y : +|[fA/a]fT|+ = f(·); return y : ↑+|[fA/a]fT|+]]$.
      First, let us rewrite the polarized types by definition and \cref{lemma:pol-subst}:
      $[[+|fΓ|+ ; +|fΦ|+ ⊢ let f : ↓∀α⁺.↑+|fT|+ = c; let y : [+|fA|+/α⁺]+|fT|+ = f(·); return y : ↑[+|fA|+/α⁺]+|fT|+]]$.
      Then we prove this judgment gradually by applying the corresponding rules of \fexists:
      \begin{enumerate}
        \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+, y : [+|fA|+/α⁺]+|fT|+ ⊢ return y : ↑[+|fA|+/α⁺]+|fT|+]]$
          holds by \ruleref{\ottdruleDTVarLabel} and \ruleref{\ottdruleDTReturnLabel}.
        \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ let y : [+|fA|+/α⁺]+|fT|+ = f(·); return y : ↑[+|fA|+/α⁺]+|fT|+]]$
          holds by \ruleref{\ottdruleDTAppLetAnnLabel}. Let us show that the 
          required premises hold. 
          \begin{enumerate}
            \item $[[+|fΓ|+ ⊢ [+|fA|+/α⁺]+|fT|+]]$ holds since $[[+|fΓ|+ ⊢ +|fA|+]]$ and $[[+|fΓ|+ ⊢ +|fT|+]]$ by \cref{lemma:wf-subst}.
            \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ f : ↓∀α⁺.↑+|fT|+]]$ by \ruleref{\ottdruleDTVarLabel}.
            \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ ∀α⁺.↑+|fT|+ ● · ⇒> ∀α⁺.↑+|fT|+]]$ by \ruleref{\ottdruleDTEmptyAppLabel}.
            \item $[[+|fΓ|+  ⊢ ∀α⁺.↑+|fT|+ ≤ ↑[+|fA|+/α⁺]+|fT|+ ]]$ by \ruleref{\ottdruleDOneForallLabel}:
              $[[+|fΓ|+  ⊢ +|fA|+/α⁺ : {α⁺}]]$ since $[[+|fΓ|+  ⊢ +|fA|+]]$, as noted above.
            \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+, y : [+|fA|+/α⁺]+|fT|+ ⊢ return y : ↑[+|fA|+/α⁺]+|fT|+]]$ as noted above.
          \end{enumerate}
        \item $[[+|fΓ|+ ; +|fΦ|+  ⊢ let f : ↓∀α⁺.↑+|fT|+ = c; let y : [+|fA|+/α⁺]+|fT|+ = f(·); return y : ↑[+|fA|+/α⁺]+|fT|+]]$
          holds by \ruleref{\ottdruleDTCVarLetLabel}. Let us show that the
          required premises hold.
          \begin{enumerate}
            \item $[[+|fΓ|+ ⊢ ↓∀α⁺.↑+|fT|+]]$ by \cref{lemma:typ-pol-wf-completeness}.
            \item $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑↓∀α⁺.↑+|fT|+]]$ by the induction hypothesis, as noted above.
            \item $[[+|fΓ|+  ⊢ ↑↓∀α⁺.↑+|fT|+ ≤ ↑↓∀α⁺.↑+|fT|+]]$ by reflexivity (\cref{lemma:subtyping-reflexivity}).
            \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ let y : [+|fA|+/α⁺]+|fT|+ = f(·); return y : ↑[+|fA|+/α⁺]+|fT|+]]$ as noted above.
          \end{enumerate}
      \end{enumerate}
  \end{caseof}
\end{proof}
